---
title: "03_recipe_alldates"
author: "Kathryn James McGraw"
date: "2026-04-29"
output: html_document
---

## Goal
Build a complete, reproducible workflow that could realistically be used to generate predictions in practice. You do not need to actually deploy to a live service unless you want to—but you must produce a **deployable artifact** (e.g., saved model + documented input schema + reproducible pipeline).

```{r packages}
# ACCRE environment
# module load StdEnv/2023 StdEnvACCRE/2023 R/4.5.0

library(tidyverse)
library(tidymodels)
library(janitor)
library(data.table)
library(stringr)
library(arrow)
library(dplyr)
library(vip)
library(purrr)
library(readr)
library(recipes)
library(xgboost)
library(finetune)
```

```{r load data produced by file 01_Data_KJM}
data_dir <- "/v5000/data/goldring_group/LPO7550/Sample"
output_dir <- "/v5000/data/goldring_group/LPO7550/Results/SimAnnealResults"

df <- read_parquet(
  file.path(data_dir, "ExiterFullSample.parquet")
) %>%
  clean_names()
```

```{r mutating outcome}
df <- df %>%
  mutate(
    returner = factor(
      if_else(returner == 1, "Returner", "NotReturner"),
      levels = c("NotReturner", "Returner")
    )
  )
```

## What to do

### 2) Data and preprocessing pipeline (reproducible)
Create a single workflow/pipeline that handles:
- Train/test split (and CV on training set as needed)
- Missing data strategy
- Encoding for categoricals
- Scaling/transformations (if needed)
- Feature engineering (if used)
- Leakage prevention (make sure no post-outcome features)

Document:
- Final feature list
- Input schema (variable names, types, allowable ranges/categories)

```{r recoding post-exit observations to prevent leakage}

setDT(df) #convert to data.table for faster processing

stems <- c(
  "career_stage", "ed_exp", "move_org_count", "field", "degree",
  "ed_field_transition", "mapped_role", "startdate_parsed", "enddate_effective",
  "date_calc_valid_flag", "education_role", "teacher_role", "school_leader_soc_role",
  "sch_lead_exp", "teaching_exp", "move_field", "move_field_count", "move_geo",
  "move_geo_count", "move_org", "role_length", "yrs_obs", "move_field_flag",
  "move_geo_flag", "move_org_flag", "role_field", "leader_role_label", "ever_role2_4",
  "start_salary", "additional_compensation", "total_compensation", "end_salary",
  "salary", "location_raw", "metro_area", "city", "country", "msa",
  "position_number", "region", "seniority", "state", "company_cleaned",
  "title_translated", "soc_cleaned", "company_cleaned_is_ed_org_match",
  "role_field_label"
)

pattern <- paste0("^(", paste(stems, collapse = "|"), ")_\\d+$")

target_cols <- names(df)[stringr::str_detect(names(df), pattern)]

for (col in target_cols) {
  col_num <- as.integer(str_extract(col, "\\d+$"))
  df[!is.na(exit_index) & exit_index < col_num, (col) := NA]
}

df <- as_tibble(df) #convert back for tidyverse below
```

```{r input schema}

input_schema <- df %>%
  summarise(across(everything(), list), .groups = "drop") %>%
  pivot_longer(everything(), names_to = "variable", values_to = "values") %>%
  mutate(
    type = map_chr(values, ~ class(.x)[1]),
    n_unique = map_int(values, ~ dplyr::n_distinct(.x, na.rm = TRUE)),
    min = map_dbl(values, ~ if (is.numeric(.x)) min(.x, na.rm = TRUE) else NA_real_),
    max = map_dbl(values, ~ if (is.numeric(.x)) max(.x, na.rm = TRUE) else NA_real_),
    example_values = map_chr(
      values,
      ~ {
        vals <- unique(.x)
        vals <- vals[!is.na(vals)]
        paste(head(vals, 5), collapse = ", ")
      }
    )
  ) %>%
  select(variable, type, n_unique, min, max, example_values)

write_csv(
  input_schema,
  file.path(output_dir, "input_schema.csv")
)
```

```{r split}
set.seed(123) #seed for reproducibility

df_split <- initial_split(df)

train_cal <- training(df_split)
test  <- testing(df_split)

# Further split training into proper train (75%) and calibration (25%)
cal_split <- initial_split(train_cal, prop = 0.75)
cal_train  <- training(cal_split)
cal_test   <- testing(cal_split)
```

```{r recipe}
formula <- as.formula("returner ~ .")

xgb_recipe_alldates <- recipe(formula, cal_train) %>%
  update_role(user_id, new_role = "id") %>%
  update_role(returner, new_role = "outcome") %>%
  step_date(all_date_predictors(), features = c("year")) %>%
  step_rm(starts_with("position_id")) %>% #meaningless; removing numeric IDs assigned to each position when data were at position-level
  step_rm(ever_entrant) %>% #ever_entrant contains the outcome, as it flagged moving from non-ed role back to ed as an entrance; removing to prevent leakage
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors(), new_level = ".missing") %>%
  step_zv(all_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>% #repeating because continuing to get "following columns have zero variance so scaling cannot be used" error message after the reordering we discussed
  step_normalize(all_numeric_predictors())
```

```{r final feature list}
xgb_prep_alldates <- prep(xgb_recipe_alldates, training = cal_train)

final_features_alldates <- juice(xgb_prep_alldates)

feature_list_alldates <- tibble(
  feature_name = names(final_features_alldates),
  type = map_chr(final_features, ~ class(.x)[1])
)

write_csv(
  feature_list_alldates,
  file.path(output_dir, "final_feature_list_alldates.csv")
)
```