---
title: "04_model"
author: "Kathryn James McGraw"
date: "2026-04-28"
output: html_document
params:
  recipe_file: "03_recipe_full.rmd"
  recipe_object: "xgb_recipe"
  suffix: "full"
---

```{r packages}
# ACCRE environment
# module load StdEnv/2023 StdEnvACCRE/2023 R/4.5.0

library(tidyverse)
library(tidymodels)
library(janitor)
library(data.table)
library(stringr)
library(arrow)
library(dplyr)
library(vip)
library(purrr)
library(readr)
library(recipes)
library(xgboost)
library(finetune)
library(knitr)
library(shiny)
library(ggplot2)
```

```{r setup for variable-omitting recipe permutations}
options(knitr.duplicate.label = "allow")

# Run selected recipe Rmd first
recipe_script <- tempfile(fileext = ".R")
knitr::purl(params$recipe_file, output = recipe_script, documentation = 0)
source(recipe_script)

# Standardize recipe name so rest of 04_model can stay mostly unchanged
xgb_recipe <- get(params$recipe_object)

# Suffix all outputs so files do not overwrite each other
model_suffix <- params$suffix
```


### 3) Model selection + tuning plan
- Define candidate model(s) and the final chosen model.
- Use a consistent tuning/evaluation procedure (e.g., CV on training set).
- Justify the choice of final model (performance + simplicity + interpretability + stability).

```{r hyperparameters to tune}
#setting model to tune mtry and depth
xgb_spec <- boost_tree(
  trees = 100, 
  tree_depth   = tune(),   # controls tree complexity
  min_n        = tune(),   # minimum observations per node
  loss_reduction = tune(), # γ: minimum gain to accept a split - REGULARIZATION; if split doesn't provide sizable upgrade to fit, downgraded towards 0
  sample_size  = tune(),   # row subsampling fraction (proportion of fold retained in tree)
  mtry         = tune(),   # column subsampling per split (variables retained)
  learn_rate   = tune()    # η: step size - how much error we're adding each time working within the residual - "help the model wander through the possible parameter space to not zero in too quickly on what is in the training set" to better handle incoming data. Hence XGB being a substantial improvement over RF.
) %>%
  set_mode("classification") %>%
  set_engine("xgboost")
```

```{r workflow}
#defining forest workflow
xgb_wf <- workflow() %>%
  add_recipe(xgb_recipe) %>% #robustness checks: xgb_recipe_nopostexit or xgb_recipe_noexityear
  add_model(xgb_spec)

#defining resamples
df_rs <- vfold_cv(cal_train, v = 20)

#set metrics
hs_metrics<-metric_set(roc_auc)

# define simulated annealing
sim_anneal_control<-control_sim_anneal(cooling_coef=.02,  
                                        verbose = FALSE,
                                        save_history = FALSE)
```

```{r tune parameters}
fit_model<- FALSE #don't need to retune hyperparameters for permutation

if(fit_model){
xgb_tune_fit <- xgb_wf %>%
    tune_sim_anneal(df_rs,
                    initial = 4,
                    metrics=hs_metrics,
                    control=sim_anneal_control)
save(xgb_tune_fit, file = file.path(output_dir, "xgb_tune_fit_proj.Rdata"))
} else{
  load(file.path(output_dir, "xgb_tune_fit_proj.Rdata"))
}
```

```{r check fit}
autoplot(xgb_tune_fit)

#inspecting fit
xgb_tune_fit %>%
  collect_metrics() %>%
  group_by(min_n,mtry)%>%
  filter(.metric == "roc_auc") %>%
  select(mean, min_n, mtry) %>%
  pivot_longer(min_n:mtry,
    values_to = "value",
    names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "AUC")

show_best_tbl <- show_best(xgb_tune_fit, metric = "roc_auc", n = 10)
write_csv(
  show_best_tbl,
  file.path(output_dir, paste0("xgb_show_best_roc_auc_", model_suffix, ".csv"))
) #save show_best output to csv (truncated in knit html)

#finalize model with best parameters
best_auc <- select_best(xgb_tune_fit,metric = "roc_auc")

final_xgb <- finalize_model(
  xgb_spec,
  best_auc)

best_auc
```

```{r final model}
#defining workflow
final_wf <- workflow() %>%
  add_recipe(xgb_recipe) %>%
  add_model(final_xgb)
```

### 4) Final evaluation and model quality checks
Provide:
- Final test-set performance with your primary metrics
- Calibration check (especially for classification) if applicable
- Robustness checks:
  - Sensitivity to preprocessing choices
  - Performance by key subgroups (if ethically and legally appropriate and variables are available)
  - Error analysis on worst-performing segment(s)
  
```{r calibration check}
# fit on cal_train for calibration
cal_fit <- final_wf %>%
  fit(data = cal_train)

# Predict probabilities on the calibration set
cal_preds <- augment(cal_fit, new_data=cal_test) %>%
  mutate(
    true_class_prob = if_else(
      returner == "Returner",
      .pred_Returner,
      .pred_NotReturner
    ),
    score = 1 - true_class_prob
  )

# Nonconformity score: 1 - predicted probability of the *true* class
cal_scores <- cal_preds %>%
  mutate(
    true_class_prob = if_else(returner == "Returner", .pred_Returner, .pred_NotReturner),
    score           = 1 - true_class_prob
  )

cal_scores %>%
  select(returner, .pred_Returner, .pred_NotReturner, score) %>%
  head(10)
```

```{r calibration check quantile_threshold}
alpha <- 0.10  # target: 90% coverage

n <- nrow(cal_scores)

# Finite-sample corrected quantile
q_level <- ceiling((n + 1) * (1 - alpha)) / n
q_level <- min(q_level, 1)  # cap at 1

q_hat <- quantile(cal_scores$score, probs = q_level)

cat("Calibration set size:", n, "\n")
cat("Quantile level used:", round(q_level, 4), "\n")
cat("Score threshold (q_hat):", round(q_hat, 4), "\n")


```

```{r calibration check score_distribution}
ggplot(cal_scores, aes(x = score)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.85) +
  geom_vline(xintercept = q_hat, color = "firebrick", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = q_hat + 0.02, y = Inf, vjust = 2,
           label = paste0("q̂ = ", round(q_hat, 3)), color = "firebrick", size = 4) +
  labs(
    title    = "Distribution of calibration nonconformity scores",
    subtitle = paste0("Scores above the dashed line exceed the ", round(1 - alpha, 2) * 100, "% threshold"),
    x        = "Nonconformity score  (1 − predicted probability of true class)",
    y        = "Count"
  ) +
  theme_minimal()
```

```{r}
#checking on testing
final_fit <- final_wf %>%
  fit(data = train_cal)

test_preds <- augment(final_fit, new_data = test)
```

```{r test conformal prediction intervals}
test_conformal <- test_preds %>%
  mutate(
    score_returner     = 1 - .pred_Returner,
    score_not_returner = 1 - .pred_NotReturner,
    pred_set = case_when(
      score_returner <= q_hat & score_not_returner <= q_hat ~ "both",
      score_returner <= q_hat ~ "Returner",
      score_not_returner <= q_hat ~ "NotReturner",
      TRUE ~ "empty"
    )
  )

efficiency <- test_conformal %>%
  mutate(
    set_size = case_when(
      pred_set == "{0,1}" ~ 2,
      pred_set %in% c("{0}", "{1}") ~ 1,
      TRUE ~ 0
    )
  ) %>%
  summarise(
    avg_set_size = mean(set_size),
    pct_singleton = mean(set_size == 1)
  )

efficiency
```

```{r conformal preds heatmap}
test_conformal %>%
  count(returner, pred_set) %>%
  ggplot(aes(x = pred_set, y = returner, fill = n)) +
  geom_tile() +
  geom_text(aes(label = n), color = "white") +
  scale_fill_viridis_c() +
  labs(
    title = "Conformal Prediction Set Distribution",
    x = "Prediction Set",
    y = "True Class"
  ) +
  theme_minimal()
```


```{r assess performance}
#confusion matrix
cm_table <- conf_mat(test_preds, truth = returner, estimate = .pred_class)
cm_table

# heat-coded confusion matrix figure
cm_table %>%
  autoplot(type = "heatmap") +
  labs(
    title = "Confusion Matrix",
    x = "Predicted class",
    y = "True class"
  ) +
  theme_minimal()

#accuracy, precision, recall, etc.
metric_set(accuracy, precision, recall, spec)(test_preds, truth = returner, estimate = .pred_class, event_level = "second")

#roc_auc
roc_auc(test_preds, truth = returner, .pred_Returner, event_level = "second")
```

### 5) Interpretability and communication
Include one or more:
- Feature importance / coefficients
- Example predictions (2–5 cases) with explanation
- Model limitations: where it should *not* be used

  
```{r extract workflow}
#extracting trained workflow for deployment
saveRDS(final_fit, file.path(output_dir, paste0("xgb_final_workflow_", model_suffix, ".rds")))

xgb_engine_fit <- final_fit %>%
  extract_fit_parsnip() #saving for VIP plots below
```

```{r}
vip(xgb_engine_fit, geom = "point") +
  geom_point(color = "steelblue") +
  labs(title = "XGBoost variable importance (Gain)") +
  theme_minimal()
```

```{r}
vip(xgb_engine_fit, geom = "point", type="cover") +
  geom_point(color = "steelblue") +
  labs(title = "XGBoost variable importance (Coverage)") +
  theme_minimal()
```

```{r}
vip(xgb_engine_fit, geom = "point", type="frequency") +
  geom_point(color = "steelblue") +
  labs(title = "XGBoost variable importance (Frequency)") +
  theme_minimal()
```

```{r sample predictions}
set.seed(2026)

example_preds <- test_preds %>%
  mutate(.row = row_number()) %>%
  select(.row, returner, .pred_class, starts_with(".pred_")) %>%
  slice_sample(n = 5)

example_cases <- test %>%
  mutate(.row = row_number()) %>%
  inner_join(example_preds, by = c(".row", "returner")) #grabbing underlying features for those 5 cases

write_csv(
  example_preds,
  file.path(output_dir, paste0("xgb_test_example_predictions_", model_suffix, ".csv"))
)
```