---
title: "06_Deployment"
author: "Kathryn James McGraw"
date: "2026-04-22"
output: html_document
---

This deployment script was generated with *lots* of help from Google and ChatGPT. 

To deploy the model, the user must have LinkedIn data (or appropriately formatted resume data) and have already generated intermediate variables, like binary flags for whether or not a role was in teaching or school leadership specifically or education overall. Depending on the data source the user has, file 01 (data wrangling) may be needed to reshape long position- and education event-level data to worker-level data. If the user's data are already worker-level, file 01 may be unnecessary. Once data are worker-level, the model can be deployed using the below script.

```{r loading packages}
library(tidymodels)
library(readr)
library(dplyr)
library(janitor)
```

```{r paths}
input_dir <- "PATH"
output_dir <- "PATH"
```

```{r function for predictions}
predict_xgb_model <- function(
  model_path,
  new_data,
  output_path = NULL
) {
  
  # 1) Load model
  model <- readRDS(model_path)
  
  # 2) Basic schema alignment
  new_data <- new_data %>%
    clean_names()
  
  # 3) Generate predictions
  preds_class <- predict(model, new_data, type = "class")
  preds_prob  <- predict(model, new_data, type = "prob")
  
  # 4) Combine outputs
  results <- new_data %>%
    select(user_id) %>% #if using data without numeric user IDs, delete this line.
    bind_cols(preds_class, preds_prob)
    
  # 5) Save to CSV
  if (!is.null(output_path)) {
    write_csv(results, output_path)
  }
  
  return(results)
}
```

```{r generating predictions}
df_new <- arrow::read_parquet(file.path(input_dir, "FILE_NAME.parquet"))

preds <- predict_xgb_model(
  model_path = file.path(input_dir, "xgb_final_workflow.rds"),
  new_data = df_new,
  output_path = file.path(output_dir, "new_predictions.csv")
)
```
